---
title: "Mono Lake Water Levels"
author: "David Segan"
date: "10/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

Visualization of Mono Lake water levels over time and the effect of diversions to meet the water demands of Los Angeles.

```{r, include=FALSE}
# ----------------
# David Segan
# ESM 206 Assignment 3 - Task 1
# Mono Lake Levels
# 10/20/19
# --------------

# --------------
# Data sources: 
# 1850-1912 from Stine, Scott (based on occasional observations and San Francisco precipitation). 
# 1912-1979 from LADWP and USGS compilations. 
# 1979-present from Los Angeles Aqueduct Daily Reports, and observations by the Mono Lake Committee. Compiled by the Mono Lake Committee.	
# ----------------

# ----------
# Attach libraries
# ---------

library(tidyverse)
library(janitor)
library(tidyr)

# ----------
# Read in mono lake data
# ---------

mono_lake <- read_csv("Mono Lake Annual Levels.csv")

# -----------
# Explore dataset
# -----------

names(mono_lake)
summary(mono_lake)

# -----------------
# Skip first 5 rows when reading in and tidy up data
# ------------------

mono_lake_2 <- read_csv("Mono Lake Annual Levels.csv", skip = 5)

mono_lake_tidy <- mono_lake_2 %>% 
  clean_names() %>% 
  mutate(land_bridges = 6377) %>% 
  mutate(brine_shrimp_decrease = 6360) %>% 
  mutate(acute_impacts_shrimp_decrease = 6350)
  

# --------------
# Graph mono lake water level over time using annotations to indicate key water levels and key historical events
# --------------

mono_lake_graph <- 
  ggplot(mono_lake_tidy, 
         aes(y = lake_level_feet_above_sea_level,
             x = year)) +
  geom_line(aes(y = lake_level_feet_above_sea_level, 
                color = vertical_change), 
            size = 2,
            show.legend = FALSE) +
  scale_color_gradientn(colors = c("red","lightblue","blue", "darkblue")) +
  geom_line(aes(y = stable_level),
            color = "black",
            linetype = "dashed") +
  annotate(geom = "text", 
           x = 1900,
           y = 6394,
           label = "Stable water level",
           size = 3.5) + 
  geom_line(aes(y = land_bridges), 
            color = "black",
            linetype = "dashed") + 
  geom_line(aes(y = brine_shrimp_decrease),
            color = "black",
            linetype = "dashed") +
  annotate(geom = "text",
           x = 1974,
           y = 6425,
           label = "Diversions begin (1941)",
           size = 3.5) + 
  annotate(geom = "curve",
           x = 1953, xend = 1941,
           y = 6425, yend = 6418,
           arrow = arrow(length = unit(.1, "inches"))) + 
  annotate(geom = "segment",
           x = 1994, xend = 1994,
           y = 6398, yend = 6379,
           arrow = arrow(length = unit(.1,"inches"))) +
  annotate(geom = "text",
           x = 1993,
           y = 6403,
           label = 
"CA water resources board
orders rise to  
stable level (1994)",
           size = 3.5) +
  annotate(geom = "text",
           x = 1910,
           y = 6379,
           label = "Land bridges allow predators to access major bird nesting sites", size = 3.5) +
  annotate(geom = "text",
           x = 1900,
           y = 6362,
           label = "Bird food sources begin to decline", size = 3.5) +
  annotate("rect", xmin = 1850, xmax = 2017,
           ymin = 6360, ymax = 6377, 
           alpha = .20, fill = "red") +
  annotate("rect", xmin = 1850, xmax = 2017,
           ymin = 6377, ymax = 6392, 
           alpha = .10, fill = "red") +
  annotate("rect", xmin = 1850, xmax = 2017,
           ymin = 6355, ymax = 6360, 
           alpha = .30, fill = "red") +
  theme_light() + 
  theme(panel.grid.minor.x = element_blank(), 
        panel.grid.minor.y = element_blank(),
        legend.position = c(.85, .75)) +
  scale_x_continuous(expand = c(0, 0), 
                     limits = c(1850, 2017), 
                     breaks = seq(1850, 2017, by = 25)) +
  scale_y_continuous(expand = c(0,0),
                     limits = c(6355, 6430),
                     breaks = seq(6360, 6430, by = 20)) +
  labs(x = "Year of measurement (recorded on October 1st)", 
       y = "Lake level (feet above sea level)",
       title = "Mono lake levels from 1850 to 2017",
       color = "Annaul change in
water level (ft)")


```


```{r}
mono_lake_graph
```

Data sources: 
1850-1912 from Stine, Scott (based on occasional observations and San Francisco precipitation). 
1912-1979 from LADWP and USGS compilations. 
1979-present from Los Angeles Aqueduct Daily Reports, and observations by the Mono Lake Committee. Compiled by the Mono Lake Committee.	